---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Unzip activity monitoring data
Unzip activity monitoring data to gain access to the csv file.
```{r unzip, cache=TRUE}
    unzip("activity.zip",exdir = "DataFile")
```

## Loading and preprocessing the data
Read the data from activity.csv and store it in activity data frame and show the summary
```{r reading, cache=TRUE}
    activity <- read.csv("DataFile/activity.csv", stringsAsFactors=FALSE)
    str(activity)
    summary(activity)
```
## Process/transform the data (if necessary) into a format suitable for analysis
Convert date to POSIXct class using lubridate package and convert interval to hour:minute format
Assuming lubridate has been previously installed
```{r date_conversion, message=FALSE}
    library(lubridate)
    activity$date <- ymd(activity$date)
    str(activity)
```

## What is mean total number of steps taken per day?
For this part of the assignment, we ignore the missing values (NA) in the dataset.

#### 1. Calculate the total number of steps taken per day
```{r daily_total,message=FALSE}
require(dplyr)
total_day <- activity %>% group_by(date) %>%summarise(total_steps=sum(steps,na.rm=TRUE),na=mean(is.na(steps))) %>% print
```
#### 2. Make a histogram of the total number of steps taken each day
```{r histogram}
total_day <- filter(total_day, na < 1)
hist(total_day$total_steps,col="blue",breaks="Sturges",main="Total steps per day",xlab="Steps Per Day")
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r summary}
mean_steps <- mean(total_day$total_steps,na.rm=TRUE)
median_steps <- median(total_day$total_steps,na.rm=TRUE)
```
The Mean and Median of the total number of steps taken per day are `r round(mean_steps,2)` steps and `r median_steps` steps respectively.

## What is the average daily activity pattern?

#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r daily, fig.width=12}
library(dplyr,quietly = TRUE)
daily_patterns <- activity %>% group_by(interval) %>% summarise(average=mean(steps,na.rm=TRUE))
plot(x = 1:nrow(daily_patterns),y = daily_patterns$average,type = "l",
     col = "blue", xaxt = "n",xlab="Intervals", 
     ylab = "Average for given interval across all days")
axis(1,labels=daily_patterns$interval[seq(1,288,12)],
     at = seq_along(daily_patterns$interval)[seq(1,288,12)])
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r daily2}
max_numb_steps_interval <- filter(daily_patterns,average==max(average))
```

Interval **"`r max_numb_steps_interval$interval`"** contains on average the maximum number of steps (**`r round(max_numb_steps_interval$average,2)` steps**).

## Imputing missing values
There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing}
na_number <- sum(is.na(activity$steps))
na_number
percentage_na <- mean(is.na(activity$steps))
percentage_na
```
Total number of missing values in the dataset is **`r na_number` ** which is **`r round(percentage_na*100,1)`** % of the total observations.

#### 2. Devise a strategy for filling in all of the missing values in the dataset
The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r strategy}
  mean(activity$steps,na.rm = T)
```

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r na_missing, cache=TRUE}
without_NAs <- numeric(nrow(activity))
for (i in 1:nrow(activity))
{
        if (is.na(activity[i,"steps"])==TRUE)
            {
                    without_NAs[i]<-filter(daily_patterns,interval==activity[i,"interval"]) %>% select(average)
            } 
        else
            {
                    without_NAs[i]<-activity[i,"steps"]
            }
                    
}
activity_without_NAs<-mutate(activity,steps_no_NAs=without_NAs)
head(activity_without_NAs)
```

#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r histogram_no_NAs}
total_day_noNAs <- activity_without_NAs %>% mutate(steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(date) %>% summarise(total_steps=sum(steps_no_NAs))
hist(total_day_noNAs$total_steps,col="red",breaks="Sturges",main="Total steps per day",xlab="Steps per day")
abline(v=median(total_day$total_steps),lty=3, lwd=2, col="black")
legend(legend="median","topright",lty=3,lwd=2,bty = "n")
```

```{r summary_no_NAs}

summary(total_day_noNAs$total_steps)
```

The impact of imputing missing data on the estimates of the total daily number of steps is that, the mean increased while median decreased,compared to estimates from the first part (ingoring missing values). Imputing missing data resulted in increase of total daily number of steps.

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r weekday}
library(lubridate)
is_weekday <-function(date){
        if(wday(date)%in%c(1,7)) result<-"weekend"
        else
                result<-"weekday"
        result
}

activity_without_NAs <- mutate(activity_without_NAs,date=ymd(date)) %>% mutate(day=sapply(date,is_weekday))

table(activity_without_NAs$day)
```

#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r weekend_comparison, cache=TRUE, fig.width=10}
library(ggplot2)
daily_patterns <- activity_without_NAs %>% mutate(day=factor(day,levels=c("weekend","weekday")),steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(interval,day) %>% summarise(average=mean(steps_no_NAs))
qplot(interval,average,data=daily_patterns,geom="line",facets=day~.)
```